Abstractions for Type-Level Programming

### 2022-04-22 article.tex
Address Dragana's comments

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
\begin{diff}
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representative for abstractions and type abstractions, respectively.
\end{diff}
Elements of $\P$ can be understood as equivalence classes for \Fm's types.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

@@

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
\begin{diff}
This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
Although $\Γ \⊢ \S \⇌ \T$ witness mutual subtyping between $\S$ and $\T$, that relation is asymmetric since its left-hand side is always a match type.
\end{diff}
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

@@

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called |Nothing|, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
\begin{diff}
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} in the presence of empty types (disjointness/subtyping exclusivity).
\end{diff}

@@

\begin{diff}
In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotation of higher-kinded types.
In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.

At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
However, it would be conceivably to lift this restriction in the future, as shown by the analysis presented in this section.

Consider the following definition of a match type |M|:
%
\begin{lstlisting}
type M[X] = X match
  case P1 => E1
  case P2 => E2
  case P3 => E3
\end{lstlisting}
%
In \Cref{sec:formalization}, we define the semantics of this type in terms of subtyping rules.
Given this particular definition of |M|, we can write a direct semantic definition of |M| with equations:
%
\begin{itemize}
 \item |M[X] = E1| \quad if |X <: P1|
 \item |M[X] = E2| \quad if |disj(X, P1)| and |X <: P2|
 \item |M[X] = E3| \quad if |disj(X, P1)| and |disj(X, P2)| and |X <: P3|
\end{itemize}
%
This equational definition is consistent under the assumptions of \Cref{lem:disjointness-subtyping-exclusivity}.

@@

In practice, it would be hard to show that |E1|, |E2|, and |E3| are co-/contravariant when type bindings get into the picture. For non-binding cases, that analysis is straightforward and analogous to the variance checks that are already in place in the compiler.
\end{diff}

@@

\begin{diff}
Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}~\citep{schinz2005compiling}, a process which preserves some types by translating them to the JVM's type systems, and removes other types by erasing them to |Object|, the JVM's top type.
Scala's type erasure directly affects pattern matching because it prevents certain patterns from being checked at runtime.
For instance, both |List[String]| and |List[Int]| erase to |List[Object]|, which makes them indistinguishable from the JVM's perspective.
This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.

In practice, the Scala compiler handles this limitation by emitting warnings for unchecked patterns.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
\end{diff}

@@

\begin{diff}
For example, consider the following function definition:
%
\begin{lstlisting}
def foo[X](x: X) = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
%
In the absence of result type annotation, Scala's type inference will first assign \lstinline!String|Int! for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
To obtain a precise result type for this method, users should annotate it explicitly:
%
\begin{lstlisting}
type Foo[X] = X match
  case Int => String
  case String => Int

>>>>

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
\begin{diff}
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representatives for abstractions and type abstractions, respectively.
\end{diff}
Elements of $\P$ can be understood as equivalence classes for \Fm's types.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

@@

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
\begin{diff}
This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$, with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
Although $\Γ \⊢ \S \⇌ \T$ witnesses mutual subtyping between $\S$ and $\T$, that relation is asymmetric, since its left-hand side is always a match type.
\end{diff}
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

@@

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called |Nothing|, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
\begin{diff}
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} (disjointness/subtyping exclusivity) in the presence of empty types.
\end{diff}

@@

\begin{diff}
In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotations of higher-kinded types.
In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.

Consider the following definition of a match type |M|:
%
\begin{lstlisting}
type M[X] = X match
  case P1 => E1
  case P2 => E2
  case P3 => E3
\end{lstlisting}
%
In \Cref{sec:formalization}, we define the semantics of this type in terms of subtyping rules.
Given this particular definition of |M|, we can write a direct semantic definition of |M| using the following equations:
%
\begin{itemize}
 \item |M[X] = E1| \quad if |X <: P1|
 \item |M[X] = E2| \quad if |disj(X, P1)| and |X <: P2|
 \item |M[X] = E3| \quad if |disj(X, P1)| and |disj(X, P2)| and |X <: P3|
\end{itemize}
%
This equational definition is consistent under the assumptions of \Cref{lem:disjointness-subtyping-exclusivity}.

@@

In practice, it would be hard to show that |E1|, |E2|, and |E3| are co-/contravariant when type bindings get into the picture. For non-binding cases, that analysis is straightforward and analogous to the variance checks that are already in place in the compiler.

At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
However, it would be conceivable to lift this restriction in the future, as shown by the analysis presented in this section.
\end{diff}

@@

\begin{diff}
Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}~\citep{schinz2005compiling}, a process which preserves some types by translating them to the JVM's type systems, and removes other types by erasing them to |Object|, the JVM's top type.
Scala's type erasure directly affects pattern matching because it prevents certain patterns from being checked at runtime.
For instance, both |List[String]| and |List[Int]| erase to |List[Object]|, which makes them indistinguishable from the JVM's perspective.
This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
Erasure hence creates an expressivity gap between pattern matching at the term and at the type level.

In practice, the Scala compiler handles this limitation by emitting warnings for unchecked patterns.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning that the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
\end{diff}

@@

\begin{diff}
For example, consider the following function definition:
%
\begin{lstlisting}
def foo[X](x: X) = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
%
In the absence of a result type annotation, Scala's type inference will first assign \lstinline!String|Int! for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
To obtain a precise result type for this method, users should annotate it explicitly:
%
\begin{lstlisting}
type Foo[X] = X match
  case Int => String
  case String => Int
